---
title: "Using raster"
author: "Lauren Hsu"
date: "3/10/2020"
output:
  md_document:
    variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## raster can be used to access the coordinate labels

Download and unzip the full spatial data files from here: https://www.angelolab.com/mibi-data

```{r echo = FALSE}
PATH <- '~/Desktop/dfci research/BIRS/mibitof/'
IMG <- 'TNBC_shareCellData/p2_labeledcellData.tiff'

setwd(PATH)

library(raster)

```

In the _TNBC_shareCellData_ there are 41 image files, each corresponding to one patient. Within those image files, there is a matrix layer containing the ID of a particular cell for that patient, which corresponds to the *cellLabelInImage*. This means that the 

The .tiff files can be saved into raster objects:
```{r}
ras_obj <- raster(IMG)
ras_obj
plot(ras_obj, main = strsplit(IMG,'/')[[1]][2])
```

The raster object can be accessed like a matrix. Each of the values in the 2000 by 2000 matrix indicates the cell lable for that location in the image. 

For example, if we zoom into the pixels in the middle of the image, we can see that cell 2395 is in the middle:

```{r}
ras_obj[1500:1510,1500:1510]
```

The raster object can also be converted easily into a matrix:

```{r}
space_mat <- as.matrix(ras_obj)
space_mat[1530:1510,1530:1510]
```

We can see that moving a bit up and to the right brings us to the border between cell 2395, cell 2423, and some background noise (0; 1)

